Inefficient Concessions and Mediation

### 2023-03-06 draft_narrow.tex
Update draft_narrow.tex
Working to broaden the language in the introduction to incorporate application to non-military domains.

In the course of a military conflict, settlement negotiations often occur simultaneously with continued fighting. If settlement can be achieved, lives are saved and destruction averted. Third-party mediators often participate in such negotiations, but the literature remains divided regarding how and under what conditions mediation can be effective (e.g., Bercovitch 2000; Greig 2005a; Beardsley et al. 2006; Duursma 2020).  This paper sheds new light on the productive role mediators can play by examining their effect on the efficiency of the concessions that combatants can offer. In our context, the efficiency of a concession is determined by how much of the material value of the concession is intact when the recipient receives it. If the party who gives the concession destroys any of its material value, the concession is deemed to be less than fully efficient.

We introduce a two-sided signaling model in which concessions can provide future value to the recipient and/or serve as a signal that provides the recipient with information about the type of player the conceding state is. Using a mechanism design framework, we demonstrate that a mediator can facilitate more efficient concessions by removing uncertainty about the ability of the parties to commit to peace. Perhaps more importantly, the mediator can also sometimes facilitate peace when the negotiating parties cannot achieve it through bilateral interaction.

Two archetypal situations illustrate the significance and difficulty of efficient concession-making in war settlement.  The first is the case of disarmament and the second is the concession of control over strategically valuable territory.

@@

This paper revises our understanding of the reasons why efficient concessions are sometimes impossible to make, why inefficient concessions can sometimes be beneficial, and how mediators can both help increase the efficiency of concessions and make peace. Weapons or strategically valuable territory will not be conceded if the conceding side fears that territory or weapons may be used in future attacks against them. One way to address this concern is to destroy the future value of the concession---e.g. decommission weapons or attempt to reduce the strategic value of the territory.  Another option, and the one on which we focus in this paper, is third-party mediation. We use this model to demonstrate that a third-party mediator can facilitate negotiated settlement by increasing trust and enabling combatants to make more efficient concessions, even when there is a risk those concessions might increase the bargaining power of their opponent in the future.

While an integral part of the literature on conflict, concessions have been only partially addressed by theoretical research. The broader literature that attempts to model war has included concessions as part of bargaining, but has not explicitly considered the role of concessions in achieving peace. Schelling's \emph{Arms and Influence} (1966) modeled war with a game theoretic framework, and Powell (1999) has also defended the use of modeling in understanding war. Fearon (1996), Powell (1996a and 1996b, 2006) and others consider war to be a bargaining failure, caused by commitment problems and asymmetric information. When the assets that are the object of the bargaining confer future power, a lack of credible commitment, or trust, can create a holdup problem that prevents the optimal level of cooperation (Fearon 1996).

@@

We will approach the commitment problem through this same general understanding, but we model a party's ability to cooperate as private information. A high type is able to cooperate; a low type has incentives not to cooperate. These types map onto the commitment problem, with a high type one who is able to commit in certain scenarios. A low type, though perhaps willing, is unable to do so. A low type may posture like the high type, but lacks the political muscle or will to follow through on a peace agreement. Democratic leaders may more often be high types because they are less likely to face coups or other hostilities in reaction to short term decisions. This is not always the case, however. Mahmous Abbas may be a low type, not because he does not want to make peace, but because he cannot credibly act on behalf of the entire Palestinian population.

>>>>

In the course of a military conflict, settlement negotiations often occur simultaneously with continued fighting. This is true in non-violent disputes as well -- parties engage in costly conflict while they negotiate. If settlement can be achieved, additional costs of conflict -- including additional death and destruction -- are averted. Third-party mediators often participate in such negotiations, but the literature remains divided regarding how and under what conditions mediation can be effective (e.g., Bercovitch 2000; Greig 2005a; Beardsley et al. 2006; Duursma 2020).  This paper sheds new light on the productive role mediators can play by examining their effect on the efficiency of the concessions that conflict participants can offer. In our context, the efficiency of a concession is determined by how much of the material value of the concession is intact when the recipient receives it. If the party who gives the concession destroys any of its material value, the concession is deemed to be less than fully efficient.

We introduce a two-sided signaling model in which concessions can provide future value to the recipient and/or serve as a signal that provides the recipient with information about the type of player the conceding state is. Using a mechanism design framework, we demonstrate that a mediator can facilitate more efficient concessions by removing uncertainty about the ability of the parties to commit to a settlement. Perhaps more importantly, the mediator can also sometimes facilitate settlement when the negotiating parties cannot achieve it through bilateral interaction.

Two archetypal situations illustrate the significance and difficulty of efficient concession-making in dispute settlement.  The first is the case of disarmament and the second is the concession of control over strategically valuable territory. In latter sections of the paper we discuss the applicability of the model to cases of nonviolent conflict as well, but peacemaking provides a particularly challenging and impactful context to illustrate the utility of inefficient concessions.

@@

This paper revises our understanding of the reasons why efficient concessions are sometimes impossible to make, why inefficient concessions can sometimes be beneficial, and how mediators can both help increase the efficiency of concessions and make peace. In the war context, weapons or strategically valuable territory will not be conceded if the conceding side fears that territory or weapons may be used in future attacks against them. More broadly, countries are reluctant to make concessions that can be used against them in latter rounds of conflict or negotiation. One way to address this concern of dangerous concessions is to destroy the future value of the concession---e.g. decommission weapons or attempt to reduce the strategic value of the territory being conceded.  Another option, and the one on which we focus in this paper, is third-party mediation. We use this model to demonstrate that a third-party mediator can facilitate negotiated settlement by increasing trust and enabling disputants to make more efficient concessions, even when there is a risk those concessions might increase the bargaining power of their opponent in the future.

While an integral part of the literature on conflict, concessions have been only partially addressed by theoretical research. The broader literature that attempts to model war has included concessions as part of bargaining, but has not explicitly considered the role of concessions in achieving peace. Schelling's \emph{Arms and Influence} (1966) modeled war with a game theoretic framework, and Powell (1999) has also defended the use of modeling in understanding war. Fearon (1996), Powell (1996a and 1996b, 2006) and others consider war to be a bargaining failure, caused by commitment problems and asymmetric information. When the assets that are the object of the bargaining confer future power, a lack of credible commitment, or trust, can create a holdup problem that prevents the optimal level of cooperation (Fearon 1996). %[OUTSIDE THE WAR CONTEXT..., ???]

@@

We will approach the commitment problem through this same general understanding, but we model a party's ability to cooperate as private information. A high type is able to cooperate; a low type has incentives not to cooperate. These types map onto the commitment problem, with a high type one who is able to commit in certain scenarios. A low type, though perhaps willing, is unable to do so. A low type may posture like the high type, but lacks the political muscle or will to follow through on a peace agreement. Democratic leaders may more often be high types because they are less likely to face coups or other hostilities in reaction to short term decisions. This is not always the case, however. Mahmoud Abbas, President of the Palestinian National Authority, may be a low type with respect to his negotiations with Israel, not because he does not want to make peace, but because he cannot credibly act on behalf of the entire Palestinian population.
